---
title: "2021 Texas Power Blackout Analysis"
author: "Emma Cardoso"
date: "2023-12-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

In February 2021, three severe winter storms swept across the United States. These storms took place between February 10-11, 13-17, and 15-20. These extreme storms were especially detrimental in the state of Texas, causing a massive electricity generation failure that resulted in a loss of power for over 4.5 million homes.[^utexas]

Given the sheer volume and destruction of these storms and the devastation that ensued, this analysis aims to investigate the number of homes that lost power during the February 2021 storms. Remotely-sensed night light data for Houston, Texas on February 7, 2021 and February 16, 2021 were examined to detect differences in night light intensity, thereby aiding to identify areas that lost electric power as a result of these storms. Additionally, data from the US Census Bureau will be used in conjunction to night light data in order to address if socioeconomic factors can serve as predictors influencing communities' recovery to loss of power. 

### Goals
The goal of this analysis is to use remotely-sensed night light data to identify the number of homes in the Houston Metropolitan area that experienced power loss as a result of the February 2021 winter storms, and to investigate if socioeconomic factors, specifically median income, can serve as an indication regarding how communities' recover from loss of power. 

### Data
#### Night Lights
To analyze the effects of the February 2021 storms in Houston, Texas, remote sensing night light data from the Visible Infrared Imaging Radiometer Suite (VIIRS) onboard the Suomi satellite will be used. This data is distrbuted through Nasa's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/), where tiles are identified based upon their horizontal and vertical positioning in the grid. As our area of interest, Houston, Texas, falls on the border of tiles h08v05 and h08v06, we will use both of these tiles to extract relevant data.

An issue that arises with remote sensing data is that environmental factors, such as cloud cover, often hinder VIIRS data from being useful. While persistent cloud cover in the Houston area did affect the ability to obtain clear and useful night light data for several days during the February 2021 storm, data from 2021-02-07 (before the storm) and 2021-02-16 (after/during the storm) provided clear night light images, allowing for the visualization of the extent of power outage in Houston, Texas during this time. 

As the Houston area falls on the border of two VIIRS tiles, two tiles per data will be utilized in this investigation, all which are stored in the `VNP46A1` folder:\

-   `VNP46A1.A2021038.h08v05.001.2021039064328.h5.tif`: tile h08v05, collected on 2021-02-07  
-   `VNP46A1.A2021038.h08v06.001.2021039064329.h5.tif`: tile h08v06, collected on 2021-02-07  
-   `VNP46A1.A2021047.h08v05.001.2021048091106.h5.tif`: tile h08v05, collected on 2021-02-16  
-   `VNP46A1.A2021047.h08v06.001.2021048091105.h5.tif`: tile h08v06, collected on 2021-02-16  

#### Roads
As the concentration of cars on highways illuminate these regions during the night, highways make up a significant portion of night light data observed from remote sensing regimes. Thus, as we are interested in visualizing the difference in night lights before and after the storm as a means of identifying areas that lost electric power, it is imperative that we do not falsely identify highways with reduced traffic as areas experiencing power outages. To ensure this, areas near highways will be omitted from this analysis.

[Geofabrik's download sites](https://download.geofabrik.de/) extracts and redistributes data from the OpenStreetMap project for public download. Geofabrik was used to download a shapefile of all highways in Texas, and from this a GeoPackage (`.gpkg` file) was created that contains only the subset of roads that intersect with the Houston metropolitan area. 

- `gis_osm_roads_free_1.gpkg`

#### Houses
As we are interested in determining the number of homes that lost power during the February 2021 storms, building data was downloaded from Geofabrik and another GeoPackage containing only data on houses in the Houston area was prepared. 

- `gis_osm_buildings_a_free_1.gpkg`

#### Socioeconomic 
Though unable to readily access socioeconomic data for every home in the Houston area, we were able to obtain data regarding 2019 census tracts from the United States Census Bureau's American Community Survey. This data is an ArcGIS file geodatabase, contained in the *folder* `ACS_2019_5YR_TRACT_48.gdb`, and includes both spatial data and demographic data. 

[^utexas]: The University of Texas at Austin, Energy Institute. 2021. "The Timeline and Events of the February 2021 Texas Electric Grid Blackouts." July 2021. https://energy.utexas.edu/research/ercot-blackout-2021#:~:text=The%20Timeline%20and%20Events%20of,more%20than%204.5%20million%20homes.

credit : this analysis is based upon materials provided by Ruth Oliver, a professor within UC Santa Barbara's Bren Graduate School of Environmental Data Science. 

## Data Analysis Workflow

```{r loading, message=FALSE, warning=FALSE}
#loading in libraries ----------

library(raster)
library(sp)
library(terra)
library(dplyr)
library(sf)
library(stars)
library(ggplot2) 
library(ggspatial)
library(stringr)

```

### Finding the Number of Homes in Houston, Texas affected by the February 2021 Blackout

In order to identify the number of homes in the Houston Metropolitan area that were affected by the power outages resulting from the winter storms in February 2021, we need to: 1) identify blackout locations, and 2) identify homes within blackout locations.

#### Part 1 : Identifying blackout locations

To identify which areas in Houston, Texas experienced blackouts, we will begin by using the remotely-sensed night light data collected on 2021-02-07 and 2021-02-16 to find the change in night light intensity between these two dates. This change in night light intensity is presumably caused by the storm. 

This will be achieved by : 
- combining the two VIIRS tiles for each date into a single `stars` object, resulting in two stars objects, one for each date
- finding the change in night light intensity between these two star objects, depicting this difference in night light intensity 
- reclassifying this difference to identify blackout areas, where areas that experienced a drop greater than 200 nW cm^-2^sr^-1^ have experienced a blackout, and locations that experienced a drop less than 200 nW cm^-2^sr^-1^ will be assigned `NA`

Once the blackout regions within the original VIIRS tiles have been identified, this blackout mask will be vectorized and cropped to the region of interest, the Houston, Texas Metropolitan area. Furthermore, once we have blackout data for the specific region of interest, highway data will be used to create a buffer, identifying areas within 200m, so we can then identify areas that experienced blackouts that are *further* than 200m from Houston highways. This is important as it will ensure that highways that experienced reduced traffic during this time are not falsely categorized as regions experiencing loss of power. 


##### Combining VIIRS data into single star objects 
```{r message=FALSE, warning=FALSE, include=TRUE, label='P1 combining star data'}
#reading in night light tiles and combining each date into a singular `stars` object ----------

#reading in light tiles and turning them into rasters, and then turning them into stars
light07_1rast <- rast("./VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif")
light07_1stars <- st_as_stars(light07_1rast)

light07_2rast <- rast("./VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif")
light07_2stars <- st_as_stars(light07_2rast)

light16_1rast <- rast("./VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif")
light16_1stars <- st_as_stars(light16_1rast)

light16_2rast <- rast("./VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif")
light16_2stars <- st_as_stars(light16_2rast)



#combining tiles into a single star object for both dates ----------

stars07 <- st_mosaic(light07_1stars, light07_2stars)
#plotting to check mosaic occurred correctly
plot(stars07, main = "2021-02-07 Night Light Intensity in nW cm-2sr-1")

#plotting to check mosaic occurred correctly
stars16 <- st_mosaic(light16_1stars, light16_2stars)
plot(stars16, main = "2021-02-16 Night Light Intensity in nW cm-2sr-1") 
```

##### Finding change in nightlight itensity and creating blackout mask 

```{r, label= 'P1 change in night lights', include=TRUE}
#finding difference in night lights intensity to create a blackout mask ----------

difference <- (stars07 - stars16) #subtracting mosaic of the 16th from that of the 7th, and setting threshold to 200 to find the change in night light intensity 

difference[difference == FALSE] <- NA #assigning NA to all values that did not experience a drop in intensity greater than 200

#plotting blackout to ensure that reclassification was correctly applied
plot(difference, main = " ")
title(main = c("Areas Experiencing a Drop in Night Light",
               "Intensity Greater Than 200 nW cm-2sr-1"),
      adj = .4,
      line = 3,
      cex.main = 1)
```
##### Cropping to region of interest
Once the blackout mask is vectorized, we will crop it to the region of interst by creating a polygon of the Houston aera by using the coordinates (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29). 

```{r include=TRUE, label='P1 vectorizing mask and cropping to region of interest'}
#vectorizing blackout mask ----------

differencevector <- st_as_sf(difference) #turning difference into an sf object
blackoutvector <- st_make_valid(differencevector) #fixing any invalid geometries of sf object

#plotting vector to check that operation occured successfully
plot(blackoutvector, main = "")
title(main = c("Vectorized Depiction of Areas Experiencing\na Drop in Night Light Intensity Greater Than 200 nW cm-2sr-1"),
      adj = .4,
      line = 2.2,
      cex.main = 1)


#cropping vectorized blackout to area of interest -----------

#defining area of interest assigning a CRS -----------

houston_matrix <- cbind(
  x = c(-96.5, -96.5, -94.5, -94.5, -96.5),
  y = c(29, 30.5, 30.5, 29, 29)) #defining coordinates

houston_polygon <-st_polygon(list(houston_matrix)) #turning coordiantes into a polygon

houston_sf <- st_sfc(houston_polygon) #converting polygon into a simple feature

houston_sf <- st_set_crs(houston_sf, st_crs(blackoutvector)) #assigning the CRS of this simple feature to match CRS of blackout vector  

#checking to make sure they have the same CRS!
if (identical(st_crs(houston_sf), st_crs(blackoutvector))) {
  print("Both data frames have the same coordinate reference system.")
} else {
  print("The data frames have different coordinate reference systems.")
}


#cropping blackout vector to regions of interest ------------

houston_blackout_crop <- blackoutvector[houston_sf, , op = st_intersects] #cropping to region of interest

houstonblackout <- st_transform(houston_blackout_crop, "EPSG:3083") #reprojecting to Texas

#plotting to check that crop occured correctly and looks different than that of blackoutvector above
plot(houstonblackout, main = "")
title(main = c("Areas Experiencing Blackout in Houston, Texas between\n2021-02-07 and 2021-02-16"),
      adj = .4,
      line = 2.5,
      cex.main = 1)
```

##### Excluding highways from blackout mask
As the GeoPackage data containing information on highways also contains data for roads and other route types, we will load in the data using a SQL query to avoid reading in data that we do not use, and will only load in data for highways. The SQL query to be used is : 
`query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"`\
`highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query)`

```{r message=FALSE, warning=FALSE, include=TRUE, label='P1 excluding highways'}
#excluding highways from blackout mask ----------

#reading in data ----------
query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'" #defining query
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query, quiet = TRUE) #loading in data
houstonhighways <- st_transform(highways, "EPSG:3083") #reprojecting to EPSG:3083

#identifying areas with 200m of all highways ----------

nearhighways1 <- st_buffer(houstonhighways, dist = 200) #creating a buffer to identify areas
nearhighways <- st_union(nearhighways1) #dissolving buffers to get a singular polygon

#performing a check to ensure that the buffer correctly occurred by making sure that the area of the buffer is greater than the area of the houstonhighways
if (all(st_area(houstonhighways) < st_area(nearhighways))) {
  print("Buffer operation successful : area of buffer > area of original Houston highways")
} else {
  print("Buffer operation failed")
}



#identifying areas that experienced a blackout, being areas that are further than 200m from a highway ----------

buffered_blackouts <- st_difference(houstonblackout, nearhighways) #using st_difference to isolate areas that fall outside of the buffer

#running a check to ensure that the difference operation occurred correctly by ensuring that the area of the buffered blackout is less than the area of the original Houston blackout, indicating that the buffered region near highways was successfully omitted
if (sum(st_area(buffered_blackouts)) < sum(st_area(houstonblackout))) {
  print("Difference operation successful : area of buffered blackout < area of original Houston blackout")
} else {
  print("Difference operation failed")
}

```

#### Part 2 : Identifying homes within blackout region
Now that we have identified which areas within the Houston Metropolitan area have been affected by blackouts, the next step is to identify homes that fall within this region. This will be achieved by using home data obtained from Geofabrik, and filtering this data to our designated blackout locations.

As the GeoPackage data we will be using contains information on all buildings, not just homes, we will be using another SQL query to select for only home data, avoiding loading irrelevant building data. The SQL query to be used is : 
`SELECT *`\ 
`FROM gis_osm_buildings_a_free_1`\
`WHERE (type IS NULL AND name IS NULL)`\
`OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')`\

```{r echo=TRUE, message=FALSE, warning=FALSE, label='P2 identifying homes within blackout region'}
#loading in building data ----------

buildingquery <- "SELECT *  FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')" #defining a query to select only for residential buildings 
buildings <- st_read("data/gis_osm_buildings_a_free_1.gpkg", query = buildingquery, quiet = TRUE) #reading in data
houstonbuildings <- st_transform(buildings, "EPSG:3083") #reprojecting data to area of interest

#finding homes within blackout areas -----------

impacted_homes <- houstonbuildings[buffered_blackouts, , op = st_intersects] #using st_intersects to filter to homes within blackout areas 

#checking to make sure filter worked correctly, by plotting all homes in houston and impacted homes, to ensure that only homes affected by blackout are displayed in impacted_homes

ggplot(houstonbuildings)+
  geom_sf()+
  ggtitle("Homes in Houston Area") #plotting all homes in Houston area

ggplot(impacted_homes)+
  geom_sf()+
  ggtitle("Homes Impacted by Blackout in Houston Area") #plotting homes impacted by blackout


cat("Number of Homes Impacted by Blackout in Houston Area:", nrow(impacted_homes), "\n") #finding number of homes
```

### Investigating Socioeconomic Factors
Now that the homes affected by the February 2021 blackout in Houston, Texas have been identified, 2019 census tract data from the United States Census Bureau's American Community Survey  (ACS) will be used to investigate if socioeconomic factors can serve as an indication regarding how communities' recover from loss of power. Specifically, looking at the median income within census tracts.

This will be achieved by :
- selecting for income data, specifically median income data, within ACS data
- joining median income data with census tract geometries
- identifying which census tracts experienced blackouts by spatially joining census tract data with building data specific to homes that were determined to be impacted by blackouts in the previous workflow 

Once the census tracts that were affected by the blackout have been identified, the median incomes of census tracts impacted by the blackout versus census tracts that were not affected will be compared by createing a map of median income by census tract (designating which tracts had blackouts), and plotting the distribution of income in impacted and unimpacted tracts. 

##### Determining which census tracts experienced blackouts
```{r message=FALSE, warning=FALSE, include=TRUE, label='determining cenus tracts blackouts'}
#loading in ACS data to investigate socioeconomic data ----------

geom <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer = "ACS_2019_5YR_TRACT_48_TEXAS", quiet = TRUE) #loading in geometries
texasgeom <- st_transform(geom, "EPSG:3083") #reprojecting 

income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer = "X19_INCOME", quiet = TRUE) #loading in income data
median_income <- income[c("GEOID","B19013e1")] #getting only median income and geoid

#performing a check to ensure only information on median income and geoid information has been preserved
cat("Checking to ensure only median income and geoid information has been selected for:\n impacted_homes column names: ", names(median_income))

#determining which census tracts experienced blackouts ----------

texascensus <- left_join(texasgeom, median_income, by = c("GEOID_Data" = "GEOID")) #joining income data to census tract geometries

censusblackout <- texascensus[impacted_homes, , op = st_intersects] #spatially joining cenus tracts data with blackout homes data

#checking to see if the spatial join was successful by comparing the number of unique census track IDs, where the joined data should have less is join was successful
cat(
  paste(
    "Number of unique census tracts in texascenus:", length(unique(texascensus$NAMELSAD)), "\n",
    "Number of unique census tracts in censusblackout:", length(unique(censusblackout$NAMELSAD)), "\n",
    "censusblackout has a smaller amount of unique census tracts, indicating the spatial join was successful\n"
  )
)

census_blackout_names <- censusblackout$NAMELSAD #determining which census tracts experienced blackouts... this could be printed but the output is very long, thus I will not print it as I can find which census tracts had blackouts by looking at the NAMELSAD column of censusblackout

``` 

##### Comparing median incomes of impacted census tracts to unimpacted census tracts
```{r message=FALSE, warning=FALSE, label= 'comparing incomes of impacted and unimpacted census tracts'} 
#comparing incomes of impacted tracts to unimpacted tracts ----------

#creating a map of median income by census tracts for all of texas 
ggplot(texascensus) +
  geom_sf(aes(fill = B19013e1), color = "white", size = 0.1) +
  scale_fill_viridis_c(name = "Median Income") +
  ggtitle("Median Income by Census Tract in Texas") +
  theme_minimal() 

#retransforming CRS of houston_sf so I can do the following operation
houstonarea <- st_transform(houston_sf, "EPSG:3083")
houstoncensus <- texascensus[houstonarea,  , op = st_intersects] #creating an object of just Houston for mapping purposes

#defining centroids for census tracts that experienced a blackout for mapping purposes
blackoutcentroids <- st_centroid(censusblackout)


#creating a map of median income by census tract for Houston specifically, so we can designate areas with blackouts
ggplot(houstoncensus) + #plotting median income for the entire houston area
  geom_sf(aes(fill = B19013e1), color = "white", size = 0.1) +
  scale_fill_viridis_c(name = "Median Income") +
  ggtitle("Median Income by Census Tract in Houston, Texas") +
 #adding centroids indicating which census tracts experienced blackouts
  geom_sf(data = blackoutcentroids, aes(color = "red"), size = .5) +
  scale_color_manual(name = "", values = "red", labels = "Census Tracts Affected by Blackouts") +
  ggtitle("Median Income by Census Tract in Houston, Texas") +
  theme_minimal()+
coord_sf() + 
  annotation_scale(location = "br") +
  annotation_north_arrow(location = "br", x = .15, y= .15,
                         pad_x = unit(0.2, "in"),
                         pad_y = unit(0.2, "in"),
                         label_size = 20, 
                         face = "bold",
                         style = ggspatial::north_arrow_nautical)

#creating an object with census information for unimpacted homes ----------

#reminding myself of which dataframes contain information on impacted homes and which contain census information for all of houston, with # so that output is not displayed :

#censusblackout         #information on areas that were affected by blackouts
#houstoncensus         #information on all homes in houston area

#creating an object of census data for houston homes that were not impacted by the blackout
unimpactedcensus <- setdiff(houstoncensus, censusblackout)

#doing a check to ensure that unimpactedcensus operation was successful, by checking to ensure that the number of rows in the impacted and unimpacted datasets sums to the number of rows in houstoncensus
if (nrow(censusblackout) + nrow(unimpactedcensus) == nrow(houstoncensus)) {
  print("Number of rows in impacted and nonimpacted homes equal the number of rows in houstoncensus data, indicating the creation of the unimpacted dataset was successful.")
} else {
  print("Number of rows mismatch. Check the creation of datasets.")
}



#plot the distribution of income in impacted and unimpacted tracts ----------

ggplot() +
  geom_histogram(data = censusblackout, aes(x = B19013e1, fill = "Impacted"), alpha = 0.5, position = "identity", bins = 30) + #histogram for impacted areas 
  geom_histogram(data = unimpactedcensus, aes(x = B19013e1, fill = "Unimpacted"), alpha = 0.5, position = "identity", bins = 30) + #histogram for unimpacted areas
  scale_fill_manual(name = "Blackout Status", values = c("Impacted" = "red", "Unimpacted" = "blue")) +
  xlab("Median Income") +
  ylab("Frequency") +
  ggtitle(str_wrap("Distribution of Median Income in Census Tracts Impacted versus Unimpacted by Blackout in Houston, Texas from 2021-02-07 to 2021-02-16")) +
  theme_minimal()

```

## Results

Upon plotting the median income distribution for census tracts impacted by the February 2021 storms in Houston, Texas, versus impacted tracts, socioeconomic disparities in blackout experience emerge. While overall median income range appears similar between impacted and unimpacted tracts, the histogram reveal notable distinctions in frequencies. Impacted tracts exhibit a higher frequency of lower median incomes, particularly in the 20,000-60,000 range, than impacted tracts. However it is important to consider that this data excluded areas within 200 meters of highways, which may inadvertently omit lower income census tracts, as proximity to highways is often associated with lower-cost housing. Another limitation to consider is that the night light data used to assess blackouts spans a 9-day gap due to cloud cover issues, meaning the nuanced dynamics of power restoration may not be captured, potentially leading to inaccuracies in identifying areas with persistent outages or rapid recovery. However, overall, the results of this study indicate that census tracts with lower median incomes were disproportionately affected by the blackout.
